Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate

### 2018-03-12 paper/main.tex
Cited two papers as requested by reviewer 1

On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.

@@

As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily \cite{Gtranslate200daily2017}. Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative \cite{TranslateCommunity}. Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as \say{sexist} \cite{AlgorithmGtranslateSexist2018}. This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}. Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding procedure. The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist. This paper intends to investigate whether it does. We are optimistic that our research endeavors, if conclusive, can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.

>>>>

On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed. In the wave of algorithmic bias, some have argued for the creation of some kind of agency in the likes of the Food and Drug Administration with the sole purpose of regulating algorithmic discrimination \cite{kirkpatrick2016battling}.

@@

As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily \cite{Gtranslate200daily2017}. Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative \cite{TranslateCommunity}. Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as \say{sexist} \cite{AlgorithmGtranslateSexist2018}. This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}. Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding procedure. The researchers are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms \cite{hajian2016algorithmic} The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist. This paper intends to investigate whether it does. We are optimistic that our research endeavors, if conclusive, can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.
